---
title: "Cox regression"
---

# A manually worked out, simple example: two groups

## Load libraries
```{r}
library(tidyverse)
library(maxLik)
library(survival)
```

## Data definition

Lets enter the data in R:
```{r}
dat <- data.frame(ratID = paste0("rat", 1:5),
                  time = c(55, 50, 70, 120, 110),
                  failure = c(0, 1, 1, 0, 1),
                  group = c(0, 1, 0, 1, 1))
```

Total number of failures D:
```{r}
sum(dat$failure)
```

For convenience, rename 'group' to 'x':
```{r}
dat <- rename(dat, x = group)
dat
```

We also define an auxiliary data.frame containing events only:
```{r}
dat.events <- subset(dat, failure == 1)
```

## Partial log-likelihood function

Lets define the partial (log-)likelihood function
```{r}
pLogLik <- function(beta) {
  numerator <- with(dat.events, x * beta)
  denominator <- rep(NA_real_, length(numerator))
  for(j in seq_along(denominator)) {
    risk_set <- subset(dat, time >= dat.events[j, "time"])
    theta_j <- with(risk_set, exp(x * beta))
    denominator[j] <- log(sum(theta_j))
  }
  return(sum(numerator - denominator))
}
```

```{r}
debugonce(pLogLik)
pLogLik(3)
```


We can plot it:
```{r}
f <- Vectorize(pLogLik)
curve(f, from = -4, to = 4)
```

## Maximum partial-Likelihood estimation
```{r}
fit.ML <- maxLik(pLogLik, start = c(beta = 0))
summary(fit.ML)
```

With the `coxph` function:
```{r}
fit.cph <- coxph(Surv(time, failure) ~ x, data = dat)
confint(fit.cph)
summary(fit.cph)
```

We can reproduce the Likelihood-ratio test:
```{r}
LRT <- 2 * (fit.ML$maximum - pLogLik(0))
data.frame(LRT = LRT,
           pvalue = pchisq(LRT, df = 1, lower.tail = FALSE))
```

The Wald test is already in the `maxLik` summary output.

# A manually worked out, simple example: one continuous covariate

```{r}
dat <- data.frame(time = c(6, 7, 10, 15, 19, 25),
                  event = c(1, 0, 1, 1, 0, 1),
                  age = c(67, 62, 34, 41, 46, 28))
```

```{r}
fit <- coxph(Surv(time, event) ~ age, data = dat)
summary(fit)
```

We might express age in decades:
```{r}
dat <- mutate(dat, age_dec = age / 10)
summary(coxph(Surv(time, event) ~ age_dec, data = dat))
```

# Case study: the pharmacoSmoking dataset

## Load the data
```{r}
library(asaur)
dat <- pharmacoSmoking
head(dat)
```

## Fit the Cox model
```{r}
fit <- coxph(Surv(ttr, relapse) ~ grp + age + gender + priorAttempts, data = dat)
summary(fit)
```

We can change the contrasts as we see fit:
```{r}
dat <- mutate(dat, grp = relevel(grp, ref = "patchOnly"))
fit <- update(fit)
summary(fit)
```

# Case study: the lung cancer dataset

## Load the data
```{r}
library(survival)

dat <- lung
dat$sex <- factor(dat$sex)
```

## Nelson-AAlen estimators
```{r}
pred.NA <- survfit(Surv(time, status) ~ sex, data = dat, type = "fh")
plot(pred.NA, col = 1:2)
```

## Cox regression: predictions
```{r}
fit.cph <- coxph(Surv(time, status) ~ sex, data = dat)

pred.cph <- survfit(fit.cph, newdata = data.frame(sex = factor(1:2)),
                    type = "aalen")

plot(pred.cph, col = 1:2)
```

How does the proportional hazards assumption hold?
```{r}
plot(pred.NA, fun = "cloglog", col = 1:2)
```

```{r}
plot(pred.cph, fun = "cloglog", col = 1:2)
```


# Case Study: pharmakosmoking

## A. MODEL FOR TTR GIVEN TRT/AGE/EMPL/GENDER/RACE


```{r}
library(asaur)
d_raw <- pharmacoSmoking
```

```{r}
d_raw
```


```{r}
table(d_raw$employment,useNA="always")
```

```{r}
dat = mutate(d_raw,employment = ifelse(employment=="ft","ft","other"), 
             grp=relevel(grp,ref="patchOnly"),
             race = ifelse(race == "white","white",
                           ifelse(race == "black","black","other"))
             )
table(dat$employment)
table(dat$race)
```

```{r}
fit=coxph(Surv(ttr,relapse)~grp+age+gender+employment+race,data=dat)
```

```{r}
coef_table = broom::tidy(fit) %>% write.csv("coef_table.csv")
```

## B. Data Segmenetation

Pretend it's a new dataset
```{r}
d_new=select(dat,-ttr,-relapse)
```


```{r}
d_segmented = d_new %>% 
  mutate(risk_score = predict(fit,newdata=d_new,type="lp"))
head(d_segmented)
```

```{r}
d_segmented %>% 
  arrange(desc(risk_score)) %>%
  head(10)
```

## C. PREDICTING MEDIAN TTR AND SURV(6 MONTHS | COVARIATES)

```{r}

d0 = d_new[1,]
p_s = survfit(fit,newdata = d0)
summary(p_s)

```

```{r}
plot(p_s)

```

```{r}
p_s
```

```{r}
summary(p_s,time=180)
```

