---
title: "COVID-19 analysis based on crowdsourced data"
---

[Data source](https://docs.google.com/spreadsheets/d/e/2PACX-1vQU0SIALScXx8VXDX7yKNKWWPKE1YjFlWc6VTEVSN45CklWWf-uWmprQIyLtoPDA18tX9cFDr-aQ9S6/pubhtml)

Paper:
[Early epidemiological analysis of the coronavirus disease 2019 outbreak based on crowdsourced data: a population-level observational study](https://www.thelancet.com/journals/landig/article/PIIS2589-7500(20)30026-1/fulltext)

Date downloaded: 2020-03-11.
Latest data update: 2020-03-09 1PM EST.

DISCLAIMER:
Data quality is **very questionable**. As such, any conclusions will be very questionable too.
This is meant as a didactic exercise only. Students are invited to find better data and/or 
keep a very critical mindset while going through the results.

```{r}
library(tidyverse)
```

# Data preparation

```{r}
d_raw <- read_csv("covid19.csv",
              col_types = cols(id = 'c', case_in = 'c', age = 'd', if_onset_approximated = 'l',
                               international_traveler = 'l', domestic_traveler = 'l', traveler = 'l',
                               `visiting Wuhan` = 'l', `from Wuhan` = 'l',
                               .default = 'c'
                               ))
```

```{r}
#?
#d_raw <- d_raw[-2224,]
```


```{r}
table(d_raw$death)
```
```{r}
table(d_raw$gender, useNA = "always")
```

```{r}
as_date <- function(x) as.Date(x, format = "%m/%d/%y")
d <-
  d_raw %>%
  mutate(reporting_date = as_date(reporting_date),
         hosp_visit_date = as_date(hosp_visit_date),
         exposure_start = as_date(exposure_start),
         exposure_end = as_date(exposure_end),
         symptom_onset = as_date(symptom_onset),
         death_status = death != "0",
         death_date = as.Date(ifelse(!death %in% c("0", "1"), as.Date(death, format = "%m/%d/%y", origin = "1970-01-01"), NA), origin = "1970-01-01"),
         gender = factor(gender, levels = c("female", "male")))
```

# Binary outcome: alive/dead

## Sex impact

### Frequency tables and independence tests
```{r}
with(d, table(death_status, gender)) # absolute frequencies
with(d, round(100 * prop.table(table(death_status, gender), 2), 1)) # relative frequencies (2-frequencies per column)
with(d, chisq.test(table(death_status, gender))) # test of independence
```

H0 = probabilities equals
p-value > 0.05 --> cannot reject H0
there is no significant difference

chisq.test is an asymptotic test (tends to chisq when n -> inf)
- easy
- computationaly cheap
- widely available

- but approximate

can use Fisher exact test (no doable by hand)
- extact test

- but computationaly demanding
- memory intensive if sample size is big

```{r}
with(d, fisher.test(table(death_status, gender)))
```

odds ratio: odds man / odds woman

odds man = 6.6% / 93.4%
odds woman = 4.1% / 95.9%

what's difference in risk of death by covid19 between man and woman: 
odds ratio (1.66) with CI of (0.97-2.93)
p-value = 0.06


### Logistic regression

Yi | Gi ~ Ber(pi) , logit(pi)= B0 + B1*Gi , iid
Gi : gender of subject (0 female, 1 male)

B in R²

yi , gi observed

--> LIKELIHOOD FUNCTION:
f(y;g;B) = product[ f(yi;gi;B)  ] -> yi, gi known, we want to estimate B
           i=1;n          

B0 intercept
B1 gendermale

```{r}
summary(glm(death_status ~ gender, data = d, family = "binomial"))
```

logit(prob_of_dying_i) = B0 + B1*Gi
B0 = -3.1551, stddev = 0.22 
B1 = 0.5069, stddev = 0.26

Ho: B1 = 0, interpreted as lack of association between gender and outcome

WALD test: Bhat / SEhat ~ N(0,1)


```{r}
exp(0.5069)
```

logistic regression -> exp(B1) : odds ratio

## Age
```{r}
summary(glm(death_status ~ age, data = d, family = "binomial"))
```

age estimate > 0 --> positive relationship
p-value << 0 -> strong relationship


using age in decade:
```{r}
fit <- glm(death_status ~ I(age/10), data = d, family = "binomial")
summary(fit)
```


```{r}
exp(coef(fit)[2])
```

for a decade increase in age -> the odds ratio is greater than 2

confidence interval
```{r}
exp(confint(fit)[2,])
```
OR CI between 1.86 and 2.7

RISK OF DEATH BY COVID19 vs AGE

~2 odds per decade increase
pvalue < 2*e-16 (strongly significant)


**using a more general model to check if linear model is ok**
gam: generalized additive mdel
s(age) - smoothing(age)

```{r}
library(mgcv)
fit = gam(death_status ~ s(age), data=d, family = "binomial")
summary(fit)
```

edf : estimated degrees of freedom (1 -> linear model)

visually
```{r}
plot(fit)
```

**fitting of polynomial (quadrtic) form (with ² term)**

```{r}
fit <- glm(death_status ~ age + I(age^2), data = d, family = "binomial")
summary(fit)
```

square term not significant.


**PIECE WISE LINEAR TERMS**
```{r}
d2 = mutate(d,age_70 = ifelse(age>70,age-70,0))
```

```{r}
summary(glm(death_status ~ age + age_70, data=d2, family = "binomial"))
```

H0: B2 = 0
age_70 estimate <0 : negative contribution when age > 70
age_70 pvalue big --> no significance

## Comparing countries

```{r}
table(d$country)
```


Caveats:
1. obsolete data
2. different testing guidelines per country
3. different reporting accuracy (i.e., often only deaths from hospitals are counted)
4. ...?

```{r}
d1 <- mutate(d, country = relevel(factor(country), ref = "China")) # odds ratio compare country vs china
summary(glm(death_status ~ country, data = d1, family = "binomial"))
```

## Adjusted model

```{r}
fit <- glm(death_status ~ I(age/10) + gender + country, data = d1, family = "binomial")
summary(fit)
```

```{r}
exp(coef(fit)[2:3])
```

## Smoothing splines for age
```{r}
library(mgcv)
fit <- gam(death_status ~ age + gender + country, data = d1, family = "binomial")
```



# Continuous outcome: survival time

```{r}
library(survival)
END_OF_STUDY <- as.Date("2020-03-10", format = "%Y-%m-%d", origin = "1970-01-01")
```

From first symptom to date of data collection: March 10th, 2020
```{r}
d_surv <-
  d %>%
  filter(!is.na(symptom_onset), !is.na(death_status),
         !(is.na(death_date) & death_status)) %>%
  mutate(
    death_date = as.Date(ifelse(is.na(death_date), END_OF_STUDY, death_date), origin = "1970-01-01"),
    time = difftime(death_date, symptom_onset, units = "days")
  )
```

```{r}
fit <- survfit(Surv(time, death_status) ~ 1, data = d_surv)
plot(fit)
```

Let's zoom in:
```{r}
plot(fit, ylim = c(0.95, 1),
     xlab = "time (days)", ylab = "Survival(t)")
```

Cannot extract a median, 7 deaths out of 623 cases (~1.1%):
```{r}
fit
```

## Effects of covariates on risk of death

```{r}
summary(d_surv)
```

## Gender

```{r}
summary(coxph(Surv(time, death_status) ~ gender, data = d_surv))
```

## Country

Model does not converge: again, too few cases.
```{r}
summary(coxph(Surv(time, death_status) ~ country, data = d_surv))
```
Is risk in France really 1.8 **billion** times that in China?

## Age

This one is strong, as expected:
```{r}
summary(coxph(Surv(time, death_status) ~ I(age / 10), data = d_surv))
```

Is the effect linear, though?
```{r}
fit_age <- coxph(Surv(time, death_status) ~ pspline(age, df = 4), data = d_surv)
summary(fit_age)
```

```{r}
termplot(fit_age, col.term = 1, col.se = 1)
```

## Piecewise-linear age effect

Segments: constant 0-70; increasing after 70

```{r}
e <-
  d_surv %>%
  mutate(
    age_70p = ifelse(age <= 70, 0, age - 70)
  )
```

```{r}
fit_age_segments <- coxph(Surv(time, death_status) ~ age_70p, data = e)
summary(fit_age_segments)
```

Or more simply, before and after 70:
```{r}
fit_age_binary <- coxph(Surv(time, death_status) ~ I(age > 70), data = e)
summary(fit_age_binary)
```

Risk after 70yo is **40 times** that of people less than 70yo!
Let's see it visually:
```{r}
e$x <- factor(ifelse(e$age > 70, ">70", "<=70"), levels = c("<=70", ">70"))
plot(survfit(Surv(time, death_status) ~ I(age > 70), data = e),
     ylim = c(0.8, 1),
     col = 1:2)
```

